\documentclass[a4paper,10pt]{article}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{algorithmicx}
\usepackage[left=2cm,top=2.5cm,right=2cm,bottom=2.5cm]{geometry}

\title{Conservative averaging}
\author{ChatGPT}

\begin{document}

\maketitle

This note discusses the standard conservative remapping problem that appears in physics and numerical modeling, including finite-volume methods, electromagnetics, and computational fluid dynamics. We consider two nonuniform grids that cover the same one-dimensional physical domain, and we want to transfer cell averages from one grid to the other while preserving the total integral.

We proceed step by step.

\section{Problem setup}

Consider a one-dimensional physical domain $[x_0,x_N]$.

\begin{itemize}
  \item \textbf{Source grid} (old grid): cell edges
  $x_{1/2}, x_{3/2}, \dots, x_{N_s+1/2}$,
  with cell averages $f_i$, $i=1,\dots,N_s$.

  \item \textbf{Target grid} (new grid): cell edges
  $X_{1/2}, X_{3/2}, \dots, X_{N_t+1/2}$,
  with unknown cell averages $F_j$, $j=1,\dots,N_t$.
\end{itemize}

Both grids cover exactly the same domain:
\[
[x_0,x_N] = [x_{1/2},x_{N_s+1/2}] = [X_{1/2},X_{N_t+1/2}].
\]

\section{Conservative averaging principle}

The remapping should preserve the total integral:
\begin{equation}
\sum_i f_i\,\Delta x_i = \sum_j F_j\,\Delta X_j,
\end{equation}
where
\[
\Delta x_i = x_{i+1/2} - x_{i-1/2},
\qquad
\Delta X_j = X_{j+1/2} - X_{j-1/2}.
\]

The physical requirement is that the integrated quantity in each target cell equals the sum of the contributions from all overlapping source cells. Therefore, for each target cell $j$,
\begin{equation}
F_j = \frac{1}{\Delta X_j} \sum_i f_i\,L_{ij},
\end{equation}
where $L_{ij}$ is the overlap length between source cell $i$ and target cell $j$:
\begin{equation}
L_{ij} = \max\!\left(0,\, \min(x_{i+1/2},X_{j+1/2}) - \max(x_{i-1/2},X_{j-1/2})\right).
\end{equation}

This formula guarantees strict conservation over the full domain.

\section{Implementation (robust and exact for grids with the same domain)}

Here is a clean Python implementation. The same logic applies in C or Fortran.

\begin{verbatim}
import numpy as np

def conservative_average_same_length(xe, f, Xe):
    """
    Conservative remap (finite-volume average) from source grid xe -> Xe
    when both grids cover the same domain.

    Parameters
    ----------
    xe : array of shape (Ns+1,) - source cell edges
    f  : array of shape (Ns,)   - source cell averages
    Xe : array of shape (Nt+1,) - target cell edges

    Returns
    -------
    F  : array of shape (Nt,)   - target cell averages (conservative)
    """
    Ns = len(f)
    Nt = len(Xe) - 1
    F = np.zeros(Nt, dtype=float)

    i = 0
    for j in range(Nt):
        X1, X2 = Xe[j], Xe[j+1]
        dx_target = X2 - X1
        s = 0.0

        # Skip source cells entirely to the left of target cell j
        while i < Ns and xe[i+1] <= X1:
            i += 1

        k = i
        while k < Ns and xe[k] < X2:
            overlap = max(0.0, min(xe[k+1], X2) - max(xe[k], X1))
            if overlap > 0.0:
                s += f[k] * overlap
            k += 1

        F[j] = s / dx_target

    return F

xe = np.array([0.0, 1.0, 2.5, 4.0])   # source edges (3 cells)
f  = np.array([2.0, 3.0, 1.0])        # source cell averages
Xe = np.array([0.0, 1.5, 3.0, 4.0])   # target edges (3 cells)

F = conservative_average_same_length(xe, f, Xe)

# Check conservation
source_total = np.sum(f * np.diff(xe))
target_total = np.sum(F * np.diff(Xe))

print("Target averages:", F)
print("Source integral:", source_total)
print("Target integral:", target_total)
\end{verbatim}

\begin{itemize}
\item The algorithm is $O(N_s + N_t)$.
\item It is strictly conservative.
\item It works even when the grids are nonuniform, unaligned, and have different numbers of cells.
\end{itemize}

\section{Summary}

To conservatively transfer cell averages between two nonuniform grids that span the same physical domain, use
\begin{equation}
F_j = \frac{1}{\Delta X_j} \sum_i f_i\, \max\!\left(0,\, \min(x_{i+1/2},X_{j+1/2}) - \max(x_{i-1/2},X_{j-1/2})\right).
\end{equation}
This preserves the integrated quantity exactly, even when the grids are nonuniform, not aligned, and have different cell counts.

\section{Deriving an averaging rule from a conservation law: concrete examples}

\subsection{Example 1: 1D layered conduction and the harmonic mean}

Consider two horizontal layers in series, with current, heat, or diffusive flux flowing vertically through both. Let layer 1 have thickness $t_1$ and conductivity $\sigma_1$, and let layer 2 have thickness $t_2$ and conductivity $\sigma_2$. Apply a potential difference $\Delta V$ across the total thickness $T=t_1+t_2$.

\paragraph{Conservation principle.} In steady state, the same vertical flux $J$ passes through both layers. Locally, Ohm's law or Fourier's law gives
\[
J = \sigma\,\frac{\Delta V_{\text{layer}}}{t_{\text{layer}}}.
\]
The total potential drop is the sum of the drops across the two layers:
\begin{equation}
\Delta V = \Delta V_1 + \Delta V_2 = \frac{J t_1}{\sigma_1} + \frac{J t_2}{\sigma_2}.
\end{equation}

Define an effective conductivity $\sigma_{\mathrm{eff}}$ for the combined layer by
\begin{equation}
J = \sigma_{\mathrm{eff}}\,\frac{\Delta V}{T}.
\end{equation}
Substituting the expression for $\Delta V$ and canceling $J$ gives
\begin{equation}
\frac{1}{\sigma_{\mathrm{eff}}}
= \frac{t_1}{T}\frac{1}{\sigma_1}
+ \frac{t_2}{T}\frac{1}{\sigma_2}.
\end{equation}
Hence,
\begin{equation}
\sigma_{\mathrm{eff}}
= \frac{T}{\frac{t_1}{\sigma_1}+\frac{t_2}{\sigma_2}},
\end{equation}
which is the weighted harmonic mean, with weights given by the thickness fractions.

\subsubsection{Numerical check}

Let
\[
t_1=2\ \text{m}, \qquad t_2=3\ \text{m}, \qquad
\sigma_1=0.10\ \text{S/m}, \qquad \sigma_2=0.01\ \text{S/m}.
\]
Then
\begin{itemize}
\item $T = 2+3 = 5$,
\item $t_1/\sigma_1 = 2/0.10 = 20$,
\item $t_2/\sigma_2 = 3/0.01 = 300$,
\item $20+300=320$,
\item $\sigma_{\mathrm{eff}} = 5/320 = 0.015625\ \text{S/m}$.
\end{itemize}

If instead you average the resistivity $\rho=1/\sigma$, then the series average is arithmetic:
\begin{equation}
\rho_{\mathrm{eff}} = \frac{t_1\rho_1+t_2\rho_2}{T}.
\end{equation}
Since $1/\rho_{\mathrm{eff}}=\sigma_{\mathrm{eff}}$, both viewpoints give the same result. Indeed, with $\rho_1=10$ and $\rho_2=100$,
\[
\rho_{\mathrm{eff}} = \frac{2\cdot 10 + 3\cdot 100}{5} = \frac{320}{5}=64\ \Omega\cdot\text{m},
\]
so $1/64 = 0.015625\ \text{S/m}$.

\textbf{Interpretation.} The harmonic mean appears because the flux is conserved while the potential drops add.

\subsection{Example 2: parallel conduction and the arithmetic mean}

Now suppose the two materials are arranged side by side and experience the same potential difference, so they act in parallel. Then the total flux is the sum of the fluxes through the two branches:
\begin{equation}
J_{\mathrm{tot}} = J_1 + J_2
= \sigma_1\frac{\Delta V}{t} + \sigma_2\frac{\Delta V}{t}
= (\sigma_1+\sigma_2)\frac{\Delta V}{t}.
\end{equation}
Defining $\sigma_{\mathrm{eff}}$ by
\[
J_{\mathrm{tot}} = \sigma_{\mathrm{eff}}\frac{\Delta V}{t},
\]
gives
\begin{equation}
\sigma_{\mathrm{eff}} = \frac{1}{2}(\sigma_1+\sigma_2)
\end{equation}
for two equal-width parallel regions. More generally, one obtains a weighted arithmetic mean.

Thus, \textbf{parallel arrangements lead to arithmetic averaging}, whereas \textbf{series arrangements lead to harmonic averaging}.

\subsection{Example 3: finite-volume face conductivity on a nonuniform grid}

In finite-volume methods, or in the two-point flux approximation, one typically enforces flux continuity across a cell face. Consider two adjacent control volumes whose centers lie at distances $d_L$ and $d_R$ from the common face, with conductivities $\sigma_L$ and $\sigma_R$.

The face conductivity $\sigma_f$ is chosen so that the flux computed from the left cell, the flux computed from the right cell, and the flux through the face are all the same. Assuming linear variation between cell centers gives
\begin{equation}
\sigma_f = \frac{d_L+d_R}{\frac{d_L}{\sigma_L}+\frac{d_R}{\sigma_R}},
\end{equation}
which is a distance-weighted harmonic mean. If $d_L=d_R$, this reduces to the ordinary harmonic mean.

\subsubsection{Numerical example}

Let $d_L=d_R=1$, $\sigma_L=0.10$, and $\sigma_R=0.01$. Then
\begin{itemize}
\item numerator $=1+1=2$,
\item denominator $=1/0.10 + 1/0.01 = 10 + 100 = 110$,
\item $\sigma_f = 2/110 \approx 0.0181818\ \text{S/m}$.
\end{itemize}

This is exactly the form used in many finite-difference and finite-volume codes for face transmissibility when the coefficient is discontinuous.

\subsection{A brief note on anisotropy and tensors}

When conductivity or diffusivity is tensor-valued, the normal component across an interface is averaged like the scalar conductivity normal to the flow direction, which leads to harmonic averaging in a series configuration. In contrast, tangential components, which behave like the parallel case, are averaged arithmetically.

For a full tensor, the correct procedure is to transform to a local coordinate system aligned with the interface, average the appropriate normal and tangential components there, and then transform back.

\end{document}
